\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-05-09 doc/incertae_sedis.tex
shortened for the time being

\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/fig1/tree1-shaded}

}\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/fig1/tax-shaded}

\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/fig1/synth1-no-is-shaded}

}\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/fig1/synth1-shaded}

\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
}taxa recovers additional edges and taxon names.
Supertree construction on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
clade $C$ leads to synthesis tree (c) if \emph{incertae sedis} annotations are
ignored, but to synthesis tree (d) if \emph{incertae sedis} annotations are respected.
The tree in (d) recovers taxon names $AB$ and $A$, as well as the edge to clade $AB$
that groups $A$ and $B$. Taxon names and edges that are conditional on handling
\emph{incertae sedis}
are shaded grey.
}

@@

\subsubsection{Caveats}

When multiple I.S.
taxa have been moved to the root node of a
subproblem, they may be I.S.
over the entire subproblem, and some may
be I.S.
over others in an asymmetric manner.
Therefore, we might need
to specify additional information about the original attachment
location of the I.S.
taxa, such as their depth.
This only affects
problems that have been decomposed.

\textbf{BDR:} \emph{currently we don't actually move taxa to get them
into a subproblem.
So, is this even an issue?}

\subsection{Grafted supertree}

\emph{Question:} Does the synthesis tree contain any \emph{incertae
sedis} groups?\\ \emph{Answer:} The grafted supertree will not contain
any \emph{incertae sedis} groups.
However, when we attach pruned nodes
to a parent in the grafted supertree, we could mark such nodes
\emph{incertae sedis} if we want.

\subsection{Unpruning}

Currently the unpruner \emph{does not} require that the OTT ids are
named in the grafted solution before unpruning starts.
According to
Mark's document, he wasn't sure if such names were generated for nodes
that had an IS taxon placed inside of them, so otc-unprune-solution-
and-name-unnamed nodes throws away all the names and generates them
itself.

\textbf{BDR}: See document texttt{otcetera/doc/unprune-solution-and-name-unnamed-nodes.pdf}

The unpruner should record when unpruned nodes are \emph{incertae
sedis}.
Such nodes are unaffected by phylogenies, and so
\emph{incertae sedis} annotations for them make good sense.

@@

It would be nice to allow running conflict against the cleaned
taxonomy, though.
One way to do this would be to generated a ``placed
taxonomy'', with groups extended to include \emph{incertae sedis} taxa
that have been placed within them.
This would not require any updating
to the conflict-analysis code in the annotation step.
\begin{comment}

\subsection{Conflict service}

The current conflict service considers a group $A$ to conflict with
the taxonomy if group $A$ has an incertae sedis group $B$ placed
within it.
This doesn't affect the annotations, since taxon names are
added by the unpruner.
But it could make perfectly fine input trees
incorrectly look like they are the cause of broken taxa, if they
contain IS taxa.
Thus, it would be nice to have a modified conflict
algorithm.

\subsubsection{Current conflict algorithm}

The current conflict algorithm is pretty fast, but it works by
classifying tips into either (i) the include group or (ii) the exclude
group.
To avoid counting the exclude group for every split, we instead
count the total number of children for each node, and assume that any
children not in the include set are in the exclude set.
This is no
longer true when we have incertae sedis taxa.
I suspect that if we
want to handle incertae sedies, we'd need a third category (iii) for
``neither include group nor exclude group''.
\begin{enumerate}
 \item
Get induced trees on intersection of leaf sets \item Compute depth for
each node (nd->depth)
 \item Compute number of tips at or below each
node (nd->n\_tips)
 \item for each input tree node -> $nd$
 \begin{enumerate}
   \item skip the root \item skip monotypic \item if
its a tip then find corresponding (``terminal'') edges in synth tree
and continue \item leaves1 <- get the list of leaves in the include
group of $nd$ (in input) \item L2 <- find the total number of tips (L2
= sum {[}nd->n\_tips| nd <- leaves1{]})
   \item leaves2 <- get list of
corresponding synth leaf nodes (in synth)
   \item nodes <- find all
nodes between leaves2 and the MRCA (in synth)
   \item MRCA <- mrca of
leaves2 (in synth.
this uses the nd->depth annotation)
   \item Compute
number of tips in the include set (nd->include\_tips) below each node
in $nodes$ (in synth)
   \item if n\_include\_tips(MRCA) == n\_tips(MRCA)
then the MRCA displays $nd$
   \item if n\_include\_tips(MRCA) <
n\_tips(MRCA) then 
   \begin{itemize} 
     \item foreach node in nodes
      \begin{itemize}
         \item if (n\_include\_tips(nd) < n\_tips(nd) and
n\_include\_tips(nd) < l2)
          \begin{itemize}
            \item this is a conflict!
\end{itemize} 
           \end{itemize}
         \item if there are no conflicts, then this
is a resolved_by.

      \end{itemize}
\end{enumerate}
\end{enumerate}

\subsubsection{Modified conflict algorithm?}

This probably is outside the scope of the paper, but if we could come
up with a modified conflict algorithm, that would be nice/useful, and
probably novel.
It would also probably be slower...

\end{comment}

\section{Results}

Should we do this? We could say:  \begin{itemize} \item there are 331
incertae sedis taxa that are mentioned in input trees.
\begin{itemize}
\item what is the kingom / phylum / class / order / genus / species
\end{itemize} \item we placed 14 incertae sedis taxa inside a sister
taxon.
(``incertae sedis'', ``versus ``unplaced''?) \item we placed
$o$ incertae sedis taxa \emph{outside} all sister taxa.
\item we
confirmed $n$ incertae sedis taxa as being separate from all sister
(sampled) taxa?? \item we avoiding breaking 8 taxa that had IS taxa
placed inside them.
\item we allowed $z_{1}$ new taxa into the
synthesis tree that were incertae sedis.
\item we allowed $z_{2}$ new
taxa into the synthesis tree that are marked as extinct.
\item some
nodes have as many as $w$\emph{ incertae sedis} children, making them
unbrowseable when incertae sedis children are not excluded.
\item
$v_{1}$ input trees were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input
trees were previously excluded \emph{partially} because they are
nested within in an incertae sedis taxon.
\end{itemize} Currently the
numbers $z_{1}$ and $z_{2}$.

\section{Discussion}

One thing we could do (perhaps) that we are not currently doing, is to
have nodes marked as incertae sedis on the synth tree.
This would be
easy enough if such nodes are not affected in any way by the input
trees.
Thus, when unpruning nodes we could mark any nodes
\emph{incertae sedis} if they were marked \emph{incertae sedis} on the
taxonomy.

Secondly, I think we need to distinguish \emph{incertae sedis} taxa
that are ``unplaced'' from \emph{incertae sedis} taxa that do not
occur in any input tree.
I think that if $A$ contains child $A_{1}$
that is an input tree, and the taxon $A$ is not broken, then $A$ will
be placed, and thus any other children $A_{2},A_{3},\ldots,A_{n}$ will
also be placed, since they will be added as children of the (placed)
node $A$ by the unpruner.
This could be considered when deciding which
nodes to suppress in the tree viewer.

@@

\subsubsection{Comparison to operational definition}

The splits-based semantics for incertae sedis taxa has many desirable
properties.
In allows the use of the BUILD algorithm to assess
compatibility of split sets.
However, it also has some properties that
may not be expected.
For example, two incertae sedis siblings can be
freely interdigitated, which might not be the expected outcome.
The
current definition, also ignore ranks.
Thus if we had two incertae
sedis siblings that were genera, but other siblings were families,
then one might expect that the genus-level incertae sedis taxa could
be placed within families, but not intermixed.
Since the current
approach ignores ranks, it cannot do that.

The above definition of incertae sedis taxa does allow two incertae
sedis sister taxa to be interdigitated.
It also

Also, the splits-based semantics might not completely recover the
edit-operation semantics.
The corner cases are ...
.

\subsubsection{Intrusion into incertae sedis taxa}

In theory it could be possible for an \emph{incertae sedis} taxon $A$
to be placed within an \emph{incertae sedis} sibling $B$.
We initially
implemented this semantics.
It requires updating the recursion to
define $I(m,n)$ as the descendants of $m$ that can be access from $n$
without traversing an \emph{incertae sedis} node.

With a splits-based semantics, we cannot allow both (i) placing $A$ in
$B$ and (ii) placing $B$ in $A$ and also forbid interdigitating $A$
and $B$.

However, if we allow this behavior by reducing the exclude sets for
$A$ and $B$, then this has the unfortunate consequence that the
members of $A$ and $B$ may be inter-digitated.
In many cases,
\emph{incertae sedis} genera have siblings that are families, and the
expected semantics is that genera may be placed within their sibling
families, or may remain outside existing families.
In such cases,
inter-digitating the genera would not be expected behavior.
Thus, we
choose the semantics that \emph{incertae sedis} taxa exclude their
\emph{incertae sedis} siblings.
This is not because the other
semantics is impossible.
However, we note that use of the
interdigitating semantics allows a situation where the names for $A$
and $B$ both apply to the same node.

This semantics also complicates placement, since it is possible to
place $A$ as a sibling of $B$, and then place $B$ within $A$.
This
violates the current invariant that taxa are only ever placed within
more tipward taxa.
This makes the description of placement more
complicated, since we can no longer simply record all cases where a
parent node is not ancestor on the taxonomy.

\subsubsection{Usage in the literature}

The term \emph{incertae sedis} can be used in a number of different
ways.
For example,\emph{ }one author might describe \emph{incertae
sedis} genera that are asserted to be monophyletic, whereas another
author might describe \emph{incertae sedis} genera that may be
interdigitated.

\appendix
\section{Alternative semantics}
\subsubsection{semantics \#1b: modified to disallow inter-digitating
\emph{incertae sedis} siblings}

We additionally stipulate that it is not possible for \emph{incertae sedis} taxa
to intrude into their \emph{incertae sedis} siblings.  We thus define $\mathcal{I}(m,n)$ to
indicate descendants of $m$ that are excluded from $n$. Thus:
\begin{align*}
\mathcal{I}(m,n) & =
    \begin{cases}
        \emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
is not}\\
        \mathcal{I}(m) & \text{otherwise}
    \end{cases}
\end{align*}

We can write the exclude set of a node $n$ in terms of
the exclude set of its parent:
\begin{eqnarray*}
    \mathcal{E}(n) & = \mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in
           siblings(n)\right].
           \label{eq:exsetformone}
\end{eqnarray*}
\subsubsection{An alternative semantics that allows placing an incertae sedis group within an incertae-sedis great-uncle}

\bdrcomment{This semantics is actually designed to allow the situation in Figure \ref{fig:Placing-one-incertae}.

  The problem is (might be) as follows: if we have the tree ``((?A1,A2)A,((D)C)B)MRCA;'', where A1 is incertae sedis, then we can place A1 within D.  This semantics does not exclude A1 from B, C, or D.  So, if A is broken, then nothing prohibits A1 from intruding into B, C, D, etc.

  The purpose of this was to allow placing B as a siblign to A1 and then placing A1 within B -- \emph{if} B were incertae sedis.  But B is not incertae sedis.

  We could try to fix up this situation by adding cases to the exclude rules to say that B excludes A1 if B is not incertae sedis.  But adding cases makes things messier.  I think it also does not actually solve the problem: suppose we mark B incertae sedis, but do NOT place it as sibling to A1.  Then again if A is broken, A1 can intrude into B.  We really want to allow A1 into B only if B is placed as sibling to A1, and I'm not sure how to do that with the partial-split-based semantics.  If you see a way, then great!
}

If we want to allow placing two incertae sedis siblings within each
other, then incertae sedis siblings are never excluded.

Furthermore, consider placing $A$ as a sibling of $B$, and then
placing $B$ within $A$.
In order for this to work, $A$ must not
exclude the incertae sedis children of $B$.

In order to accomplish both of these goals, we define the descendants
$\mathcal{I}^{\prime}(n)$ that are accessible without passing through
an incertae sedis node.

\begin{align*}
    \mathcal{I^{\prime}}(n) & =
    \begin{cases}
        \emptyset & \text{if }n\in\mathcal{U}\\
        \{n\} & \text{if }n\text{ is a leaf}\\
        [\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
\end{cases}
\end{align*} We may then define the taxa excluded from $n$
as:

\begin{eqnarray*} 
\mathcal{E}(n) & =\mathcal{E}(parent(n))\cup\left[\mathcal{I}(m,n)\big|m\in siblings(n)\right].\label{eq:exsetformoneb}
\end{eqnarray*}

This semantics satisfies criteria 1,3,4, and 5 above.

\subsubsection{A third semantics}

\bdrcomment{I didn't really finish working this one out.}

OK, the idea here is that we want to avoid inter-digitation.
\emph{Incertae sedis} taxa thus exclude\emph{ incertae sedis}
siblings.
However, where the first semantics allows \emph{incertae
sedis }taxa to intrude into \emph{incertae sedis} descendants of
siblings, but not allow\emph{ incertae sedis} descendants of siblings
to intrude into \emph{incertae sedis} taxa, here we do the reverse.

OK, so the idea here is that an\emph{ incertae sedis} group excludes
\begin{itemize} \item \emph{incertae sedis} siblings \item
\emph{incertae sedis }descendants of ancestors \item but not\emph{
incertae sedis} descendants of non-incertae-sedis siblings
\end{itemize} It is also possible to allow placing $A$ as a sibling of
$B$ and then placing $B$ within $A$ if we disallow placing $A$ into
incertae sedis descendants of its siblings.
We may then allow placing
those descendants within $A$.

\begin{align*} \mathcal{I^{\prime}}(n) & =\begin{cases} \emptyset &
\text{if }n\in\mathcal{U}\\ \{n\} & \text{if }n\text{ is a leaf}\\{}
[\mathcal{I}^{\prime}(c)|c\in children(n)] & \text{otherwise.}
\end{cases} \end{align*}

We use $\mathcal{I}(m,n)$ to indicate the descendants of $m$ that are
excluded from $n$.
\begin{align*} \mathcal{I}(m,n) & =\begin{cases}
\emptyset & \text{if }n\text{ is \emph{incertae sedis}\,and }m\text{
is not}\\ \mathcal{I}(m) & \text{if }n\text{ and }m\text{ are both
}incertae\,sedis\\ \mathcal{I}^{\prime}(m) & \text{otherwise.}
\end{cases} \end{align*}

We may then define the taxa excluded from $n$ as:
\begin{align*}
\mathcal{E}(n) & =\begin{cases}
\left[\mathcal{L}-\mathcal{I}(parent(n))\right]\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{if }n\in\mathcal{U}\\
\mathcal{E}(parent(n))\cup[\mathcal{I}(s,n)|s\in siblings(n)] & \text{otherwise.}
\end{cases}
\end{align*}
I think this works... but need to double-check.

Also: can we genericize $\mathcal{E}(parent(n))$?

\section{Details \& Methods}

\subsection{Cases of incertae sedis taxa}
1.1 Some examples in OTT This paragraph pulled from intro to
otcetera/doc /handling-incertae-sedis.pdf

The problem is becoming more acute because NCBI is putting a larger
number of taxa into groups that are marked as ``classified''. For
example, when OTT 2.9 was created NCBI's clasification of the bird
family Sylvidae included a group five genera that were placed within
\url{http://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?mode=Undef&id=36270&lvl=3&lin=f&keep=1&srchmode=1&unlock||unclassified Sylviidae}.
This includes the genus Regulus. Thus in OTT 2.9 Regulus (ott ID = 3599326) is placed inside the family Silvidae (OTT ID =
259942).
The ``unclassified Sylviidae'' does not appear in OTT; instead
Regulus is flagged as ``unclassified,sibling higher'' and all of the
species within Regulus are flagged as ``unclassified inherited.''

The taxonomy (as of version 2.9) also contains 685 cases of taxa that
are flagged as both ``unclassified inherited'' and ``unclassified'' based
on {\tt grep unclass.*unclassified\_inherited taxonomy.tsv}.

We would like to stop suppressing (pruning) incertae sedis taxa, so
that groups such as Regulus can appear appear in the synthetic tree.

\bdrcomment{We get Regulus back, bu we do not end up placing it in Sylviidae
because Sylviidae is broken. It conflicts with about 20 different trees.}

\subsection{Taxonomy details}

\paragraph{Flags.}
The Open Tree's reference taxonomy is produced by a tool called
smasher that notices hints that a taxon is incertae sedis and labels
that taxon with one of five flags (see https://github.com/OpenTreeOfLife
/reference-taxonomy/wiki /Taxon-flags). This document is not concerned
with the flagging system per se, so ``incertae sedis'' will be used here
to refer to all of the flags that denote taxa with uncertain
placement.

\paragraph{Limited attachment range}
While a taxonomist may intend to use incertae sedis to
indicate a limited number of possible positions for a taxon to go in
the taxonomy, we do not retain any such details.

\paragraph{Container nodes.}
Some taxonomies have nodes with names like ``Incertae sedis (Bacteria)''.
Such nodes are called containers. They indicate that each child node is an
incertae sedis child of the container node's parent. The OTT retains container
nodes with the was\_container flag, but seems to have already moved all of the
container's children to the container's parent.

\section{Placements}
\textbf{Whale:}~ ?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott698424~{[}698424~Cetacea/orde r{]}~placed~under~ott4942423~{[}4942423~Cetotherium~/~genus{]}~<-~ott698424~{[}6 98424~Cetacea/order{]}

\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetacea~/~order{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~

-~Cetacea~broken.

\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ot t275501~{[}275501{]}~<-~ott672501~{[}672501{]}

\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Reduviidae~/~f amily{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

-~Reduviidae~broken.

\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrididae~/~family{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~

\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamioidea~/~sub family{]}~placed~under~ott534796~{[}534796{]}~<-~

\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamioidea~/~sub family~{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~

ott1074933~{[}1074933{]}~

-~Lamioidea~\emph{not}~broken

\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~o tt447802~{[}447802{]}~<-~ott23373~{[}23373{]}

-~Orobanchaceae~broken

\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~<-~ott1014 022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}571842~Pavetteae~/~tribe{ ]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]}~<-~ott1014022~{[}1014022~Rubia ceae/family{]}

\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott 6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~

\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~ under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~

\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~u nder~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~

-~Erysipelotrichaceae~broken

>>>>

\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/fig1/tree1-shaded}}
\hfill{}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/fig1/tax-shaded}}

\subfloat[Synthesis Tree (no \emph{incertae sedis})]{\includegraphics[scale=0.5]{Figures/fig1/synth1-no-is-shaded}}
\hfill{}\subfloat[Synthesis Tree (\emph{incertae sedis} aware)]{\includegraphics[scale=0.5]{Figures/fig1/synth1-shaded}}

\caption{\label{fig:Handling-incertae-sedis}Handling \emph{incertae sedis
}taxa recovers additional edges and taxon names.
Supertree construction on input tree (a) and taxonomy tree (b) with \emph{incertae sedis}
clade $C$ leads to synthesis tree (c) if \emph{incertae sedis} annotations are
ignored, but to synthesis tree (d) if \emph{incertae sedis} annotations are respected.
The tree in (d) recovers taxon names $AB$ and $A$, as well as the edge to clade $AB$
that groups $A$ and $B$. Taxon names and edges that are conditional on handling
\emph{incertae sedis} are shaded grey. \mthcomment{We should probably remove $C3$ and replace $B1$ and $B2$ with $B$
to simplify this. We could use triangle to indicate that $D$ and $B$ could be leaves or clades.}
}

@@

@@

It would be nice to allow running conflict against the cleaned
taxonomy, though.
One way to do this would be to generated a ``placed
taxonomy'', with groups extended to include \emph{incertae sedis} taxa
that have been placed within them.
This would not require any updating
to the conflict-analysis code in the annotation step.

@@
